---
title: "Financial and Health Effects of Weather Events"
author: "Charlie LeCrone"
date: "October 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This analysis will evaluate the health and financial effects of various weather events.  The data comes from........


## Load Libraries and Functions

First I will load the libraries and functions I will use in the analysis.  I'll also turn off scientific notation.  We'll be looking at a lot of large numbers in this analysis, and I prefer to see them in standard notation.

```{r, echo=T}
# load libraries

library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)

# Turn off scientific notation

options(scipen = 999)

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

## Data Processing


```{r, echo=T, cache=T}

# Check to see if data exists in working directory and download it if it is not

if(!file.exists("./repdata%2Fdata%2FStormData.csv.bz2"))
  
{
  fileURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(fileURL, destfile = "./repdata%2Fdata%2FStormData.csv.bz2")
}

# Read data file into data file d1

d1<-read.csv(bzfile("./repdata%2Fdata%2FStormData.csv.bz2"))

# Create a vector of the columns I want to keep

keep<-c(2,7,8,23,24,25,26,27,28)

d2<-d1[,keep]

# Show columns I've kept

names(d2)

# check for missing data

length(complete.cases(d2))

nrow(d2)

# trim off excess characters on dates

# convert from factor to character

d2$BGN_DATE<-as.character(d2$BGN_DATE)

# look at what dates look like

head(d2$BGN_DATE)

# trim last 8 spaces

d2$BGN_DATE<-substr(d2$BGN_DATE, 1, nchar(d2$BGN_DATE)-8)

# convert to dates

d2$BGN_DATE<-as.Date(d2$BGN_DATE, "%m/%d/%Y")

# convert date to year

d2$BGN_DATE<-year(d2$BGN_DATE)

# Update Names

names(d2)<-c("Year", "State", "Event", "Deaths", "Injuries", "PropertyDamage", "PExp", "CropDamage", "CExp")

head(d2)

# Look at levels of propergy damage multiplier

Plvllist<-levels(d2$PExp)
Clvllist<-levels(d2$CExp)

Plvllist
Clvllist

#Convert Exponent columns to character

d2$PExp<-as.character(d2$PExp)
d2$CExp<-as.character(d2$CExp)

# Convert exponent flags to numeric characters

d2<-d2 %>% 
  mutate(PExp = ifelse(PExp == "H" | PExp == "h", "2", PExp)) %>%
  mutate(PExp = ifelse(PExp == "K" | PExp == "k", "3", PExp)) %>%
  mutate(PExp = ifelse(PExp == "M" | PExp == "m", "6", PExp)) %>%
  mutate(PExp = ifelse(PExp == "B" | PExp == "b", "9", PExp)) %>%
  mutate(PExp = ifelse(PExp == "+", "0", PExp))               %>%
  mutate(CExp = ifelse(CExp == "H" | CExp == "h", "2", CExp)) %>%
  mutate(CExp = ifelse(CExp == "K" | CExp == "k", "3", CExp)) %>%
  mutate(CExp = ifelse(CExp == "M" | CExp == "m", "6", CExp)) %>%
  mutate(CExp = ifelse(CExp == "B" | CExp == "b", "9", CExp)) %>%
  mutate(CExp = ifelse(CExp == "+", "0", CExp))

# Convert numeric flags to actual numbers - creates NAs for non numerics

d2$PExp<-as.numeric(d2$PExp)
d2$CExp<-as.numeric(d2$CExp)

# Convert NAs to -1

d2 <- d2 %>%
  mutate(PExp = ifelse(is.na(PExp),-1,PExp)) %>%
  mutate(CExp = ifelse(is.na(CExp),-1,CExp))


# Convert Property Damage and Crop Damage estimates by multiplying by the exponents

d2<- d2 %>%
  mutate(PropertyDamage = ifelse(PExp == -1, 0, PropertyDamage*10^PExp)) %>%
  mutate(CropDamage = ifelse(CExp == -1, 0, CropDamage*10^CExp))

# Group by Events and total damage and health statistics

d3<-d2 %>%
  group_by(Event) %>%
  summarize(PropertyDamage=sum(PropertyDamage), CropDamage = sum (CropDamage), Deaths=sum(Deaths), Injuries=sum(Injuries))

# Add Deaths + Injuries and Crop + Property Damage

d3$TotalHealth<-with(d3, Deaths+Injuries)
d3$TotalDamage<-with(d3, PropertyDamage+CropDamage)

# Select only the columns I need for plots

keep3<-c(1,6,7)

d3<-d3[,keep3]

# Arrange on table by Health Impact and another by Damage

p1<-arrange(d3, desc(TotalHealth))

p2<-arrange(d3, desc(TotalDamage))

# Select the top 15 Health Impact Events and the top 10 Damage Events

p1<-p1[1:15,]

p2<-p2[1:15,]

# look at plotting data frames

head(p1)

head(p2)
```

## Results



```{r, echo=T}

# Create plots of top 15s for Health and Damage Impact

plot1<-ggplot(aes(x=Event, y=TotalHealth),data=p1)+geom_bar(fill="blue", stat="identity")+
  labs(x="Event", y="Deaths + Injuries", title="Top 15 Weather Events in Terms of Deaths+Injuries")+
  theme(legend.position = "none", axis.text.x = element_text(angle = 60,hjust = 1))+
  scale_x_discrete(limits=p1$Event)+scale_y_continuous(labels = comma)

plot2<-ggplot(aes(x=Event, y=TotalDamage),data=p2)+geom_bar(fill="blue", stat="identity")+
  labs(x="Event", y="Total Damage in Dollars", title="Top 15 Weather Events in Terms of Property + Crop Damage")+
  theme(legend.position = "none", axis.text.x = element_text(angle = 60,hjust = 1))+
  scale_x_discrete(limits=p2$Event)+scale_y_continuous(labels = comma)


multiplot(plot1, plot2, cols=2)
```

There is a lot of variability in how events are labeled in the data set.  Let's check to see what percentage of the total health and financial effects are included in the top 15 events for each category:

```{r, echo=T}

# Check to see what percentage of totals are represented by the top 15

TH<-sum(d3$TotalHealth)
TD<-sum(d3$TotalDamage)
PH<-sum(p1$TotalHealth)
PD<-sum(p2$TotalDamage)

PH/TH # Percent of total health impact from top 10

PD/TD # Percent of total damage from top 10
```

Finally, we'll look at how these weather effects have trended over time

```{r, echo=T}

# Create a vector of uniqe Events from both top 15 lists

keepnames<-unique(c(as.character(p1$Event), as.character(p2$Event)))

# Create a new data table for time series plot of impactful events

year<-d2

year$Event<-as.character(year$Event)

year<-subset(year, year$Event %in% keepnames)

yearplot<-year %>%
  group_by(Year) %>%
  summarize(HealthEffects=sum(Deaths+Injuries), FinancialEffects=sum(PropertyDamage+CropDamage)) %>%
  arrange(Year)

plot3<-ggplot(yearplot, aes(x=Year, y=HealthEffects))+geom_line(col="blue")+
  labs(x="Year", y="Deaths + Injuries", title="Health Effects of Weather Events by Year")+
  geom_smooth(method = "auto")+scale_y_continuous(labels = comma)

plot4<-ggplot(yearplot, aes(x=Year, y=FinancialEffects))+geom_line(col="blue")+
  labs(x="Year", y="Property + Crop Damage", title="Financial Effects of Weather Events by Year")+
  geom_smooth(method = "auto")+scale_y_continuous(labels = comma)

multiplot(plot3, plot4, cols=2)
```

